Week 7 Principles of Spatial Analysis - Geodemographic Classification

```{r}
library(usethis)
library(tidyverse)
library(janitor)
library(ggcorrplot)
library(cluster)
library(factoextra)
library(sf)
library(tmap)
library(here)
```

Load individual csv files that we downloaded into R

```{r}
# load age data
lsoa_age <- read_csv(here::here("wk7_data", "London-LSOA-AgeGroup.csv"))

# load country of birth data
lsoa_cob <- read_csv(here::here("wk7_data", "London-LSOA-Country-of-Birth.csv"))

# load ethnicity data
lsoa_eth <- read_csv(here::here("wk7_data", "London-LSOA-Ethnicity.csv"))

# load language data
lsoa_lan <- read_csv(here::here("wk7_data", "London-LSOA-MainLanguage.csv"))
```

Since all the data are stored in long format, we need to transform it into a wide format. We will clean up column names as well.

```{r}
# clean names, pivot and clean names again
lsoa_age_wide <- lsoa_age %>% 
  janitor::clean_names() %>% 
  tidyr::pivot_wider(id_cols = "lower_layer_super_output_areas_code", names_from = "age_5_categories", values_from = "observation") %>%
  janitor::clean_names()

# inspect
View(lsoa_age_wide)
```

To account for the non-uniformity of the areal units, we further need to convert the observations to proportions

```{r}
# total observations
lsoa_age_wide <- lsoa_age_wide %>% 
  rowwise() %>% 
  mutate(age_pop = sum(across(2:6)))

# total proportions, select columns
lsoa_age_wide <- lsoa_age_wide %>% 
  mutate(across(2:6, ~./age_pop)) %>% 
  select(1:6)

# inspect
head(lsoa_age_wide)
```

Looks much better. We can do the same for the country of birth data

```{r}
# prepare country of birth data
lsoa_cob_wide <- lsoa_cob %>% 
  janitor::clean_names() %>% 
  tidyr::pivot_wider(id_cols = "lower_layer_super_output_areas_code", names_from = "country_of_birth_8_categories", values_from = "observation") %>% 
  janitor::clean_names()
(lsoa_cob_wide)

# proportions, select columns
lsoa_cob_wide <- lsoa_cob_wide %>% 
  rowwise() %>% 
  mutate(cob_pop = sum(across(2:9))) %>% 
  mutate(across(2:9, ~./cob_pop)) %>% 
  select(-2, -10)

View(lsoa_cob_wide)
```

Do the same for ethnicity and language datasets

```{r}
# ethniciy data
lsoa_eth_wide <- lsoa_eth %>% 
  janitor::clean_names() %>% 
  tidyr::pivot_wider(id_cols = "lower_layer_super_output_areas_code", names_from = "ethnic_group_20_categories", values_from = "observation") %>% 
  janitor::clean_names()

# proportions, select columns
lsoa_eth_wide <-  lsoa_eth_wide %>% 
  rowwise() %>% 
  mutate(eth_pop = sum(across(2:21))) %>% 
  mutate(across(2:21, ~./eth_pop)) %>% 
  select(-2, -22)

# language data
lsoa_lan_wide <- lsoa_lan %>% 
  janitor::clean_names() %>% 
  tidyr::pivot_wider(id_cols = "lower_layer_super_output_areas_code", names_from = "main_language_11_categories", values_from = "observation") %>% 
  janitor::clean_names()

# proportions
lsoa_lan_wide <- lsoa_lan_wide %>%
  rowwise() %>% 
  mutate(lan_pop = sum(across(2:12))) %>% 
  mutate(across(2:12, ~./lan_pop)) %>% 
  select(-2, -11, -13)
```

Examine the distribution of our variables by creating boxplots for each variable

```{r}
# wide to long
lsoa_age_wd <- lsoa_age_wide %>% 
  pivot_longer(cols = c(2:5), names_to = "agegroup", values_to = "count")

# facet age
ggplot(lsoa_age_wd, aes(y = count)) + geom_boxplot() + facet_wrap(~agegroup, ncol = 2) + theme_minimal() + ylab("")
```

```{r}
# join
lsoa_df <- lsoa_age_wide %>% 
  left_join(lsoa_cob_wide, by = "lower_layer_super_output_areas_code") %>% 
  left_join(lsoa_eth_wide, by = "lower_layer_super_output_areas_code") %>% 
  left_join(lsoa_lan_wide, by = "lower_layer_super_output_areas_code")

# calculate proportion of zeroes
zero_prop <- sapply(lsoa_df[2:41], function(x) {
  mean(x == 0)
})

# extract variables with high proportion of zeroes
idx <- which(zero_prop > 0.25)

# inspect
idx
```

```{r}
# remove variable swith high proportion of zeroes
lsoa_df <- lsoa_df %>% 
  select(-white_gypsy_or_irish_traveller,
         -any_other_uk_languages,
         -oceanic_or_australian_languages,
         -north_or_south_american_languages)
```

```{r}
# inspect variable names
names(lsoa_df)
```

```{r}
# change variable names to index to improve visualisation
lsoa_df_vis <- lsoa_df
names(lsoa_df_vis)[2:37] <- paste0("v", sprintf("%02d", 1:36))

# correlation matrix
cor_mat <- cor(lsoa_df_vis[, -1])

# correlation plot
ggcorrplot(cor_mat, outline.col = "#ffffff", tl.cex = 8, legend.title = "Correlation")
```

We define a weak correlation as lying between 0 and 0.40, moderate as between 0.41 and 0.65, and strong as between 0.66 and 0.80. However, here we decide to retain all variables to maintain representation

```{r}
# standardisation of input variables ensures that each variable contributes equally to the clustering process

# inverse hyperbolic sine on all variables
lsoa_df_vis[, -1] <- sapply(lsoa_df_vis[-1], asinh)

# range standardise across all variables
lsoa_df_vis[, -1] <- sapply(lsoa_df_vis[-1], function(x) {
  (x - min(x)/(max(x) - min(x)))
})
```

Selecting the number of clusters using the elbow method

```{r}
# elbow plot
fviz_nbclust(lsoa_df_vis[, -1], kmeans, nstart = 100, iter.max = 100, method = "wss")
```

k-means clustering

```{r}
# set seed for reproducibility
set.seed(999)

# k-means
lsoa_clus <- kmeans(lsoa_df[, -1], centers = 6, nstart = 100, iter.max = 100)
```

```{r}
# inspect
lsoa_clus
```

Visualising clusters

We can compare the global mean values of each variable with the mean values specific to each cluster

```{r}
# global means
glob_means <- colMeans(lsoa_df_vis[, -1])

# add clusters to input data
lsoa_df_vis_clustered <- cbind(lsoa_df_vis, cluster = lsoa_clus$cluster)

View(lsoa_df_vis_clustered)

# cluster means
cluster_means <- lsoa_df_vis_clustered %>% 
  group_by(cluster) %>% 
  summarise(across(2:37, mean))

# difference
cluster_diffs <- cluster_means %>% 
  mutate(across(2:37, ~. -glob_means[cur_column()]))
```

These comparisons can be visualised using a radial bar plot

```{r}
# to long format
cluster_diffs_long <- cluster_diffs %>% 
  pivot_longer(!cluster, names_to = "vars", values_to = "score")

# facet clusters
ggplot(cluster_diffs_long, aes(x = factor(vars), y = score)) + geom_bar(stat = "identity") + coord_radial(rotate.angle = TRUE, expand = FALSE) + facet_wrap(~cluster, ncol = 3) + theme_minimal() + theme(axis.text.x = element_text(size = 7)) + xlab("") + ylab("")
```

Of course, we can also map the results

```{r}
# read the spatial dataset
lsoa21 <- st_read(here::here("wk7_data", "London-LSOA-2021.gpkg"))
```

```{r}
# join
lsoa21 <- cbind(lsoa21, cluster = lsoa_clus$cluster)

# shape, polygon
tm_shape(lsoa21) +
  # specify columns, colours
  tm_polygons(
    col = "cluster",
    palette = c("#feebe2", "#fbb4b9", "#f768a1", "#c51b8a", "#7a0177"),
    border.col = "#ffffff",
    border.alpha = 0.1,
    title = "Cluster Number"
  ) +
tm_layout(legend.outside = FALSE, legend.position = c("right", "bottom"), frame = FALSE)
  )
```

This is the end of the practical.